---
title: "Eco-Driving Project"
output: html_document
date: "2026-01-20"
---
# Eco-Driving Problem statement
##Case Study
In 2026, urban transport systems face critical inefficiencies where unnecessary stop-and-go cycles at signalized intersections and aggressive driving behaviors—characterized by rapid acceleration and harsh braking—result in a 15-25% increase in fuel consumption and over 30% higher CO2 emissions. This global environmental and safety crisis primarily impacts commuters and urban residents during peak traffic hours, as vehicles repeatedly lose and rebuild kinetic energy. To mitigate these risks, data science and microscopic trajectory analysis must be deployed to optimize traffic signal synchronization and implement real-time eco-driving feedback systems, transforming granular vehicle data into actionable strategies for smoother, more sustainable traffic flow.

##Business Understanding
What: The problem is the significant fuel consumption and CO2 emissions caused by unnecessary stop-and-go driving at signalized intersections and aggressive driving behaviors like rapid acceleration and harsh braking. These behaviors increase fuel consumption by 15-25% and CO2 emissions by over 30%, while also posing higher accident risks.

Why: These issues arise because frequent acceleration from rest requires the engine to work harder, wasting the accumulated kinetic energy during subsequent braking and resulting in inefficient fuel usage and increased pollutants.

Who: This problem affects vehicle drivers, the general public (due to air quality and accident risks), transportation/planning agencies, and the environment.

Where: The problem is particularly pronounced at signalized intersections and in urban environments/congested traffic conditions.

When: The issue occurs continuously during everyday driving, but is exacerbated during peak traffic times (e.g., morning and evening rush hours) and in scenarios involving frequent speed changes.

How (Data Science & Analysis): This problem can be addressed using data science and analysis by:

Utilizing microscopic emission models that analyze granular vehicle trajectory data (speed and acceleration information) to accurately quantify emissions and consumption patterns.

Employing statistical modeling and data analysis to identify relationships between driving styles, traffic flow characteristics, and environmental impacts.

Developing data-driven eco-driving strategies or Intelligent Transport System (ITS) technologies that provide real-time information to drivers to encourage smoother, less volatile driving behaviors and minimize stops.

Implementing and evaluating changes in intersection design or traffic management systems (e.g., signal timing optimization) using data from field tests or simulations to reduce delays and improve overall traffic efficiency, thereby mitigating fuel consumption and emissions.

##Predictor Variables
- Independent variables: Idling time
- Dependent Variables: rpm_variation, harsh_braking_count, fuel_consumption, acceleration_smoothness, eco_score

## Problem statement
1. Significant fuel consumption and CO2 emissions are generated by unnecessary stop and go driving at signalized intersections, where vehicles frequently accelerate from zero, losing accumulated kinetic energy.

2. Aggressive driving behavior such as rapid acceleration and harsh braking-increase fuel consumption by 15%-25% and CO2 emissions by over 30%, in addition to posing higher accident risks.

# Focus
1. Traffic Light Information (SPaT), Green Light Optimal Speed Advisory (GLOSA)
2. Driver Coaching , Training and feedback.

# Objective
- The primary goal of this dataset is to help analyze how different driving behaviors impact fuel consumption and an overall eco-driving score, which reflects driving efficiency and environmental friendliness.

**Type of analysis being used main focus of the project to solve the real world problem
  -- State the question
  -- Response 
  -- Predictor variables

# Main Focus of Approach
To address the 2026 urban transport crisis of stop-and-go inefficiencies and aggressive driving, this approach transforms microscopic trajectory data (30,000 entries) into actionable insights for traffic optimization and real-time driver feedback.

I. Data Preparation Strategy
With 30,000 records, the focus is on maximizing the predictive signals from behavioral features like harsh_braking_count and rpm_variation.

Data Cleaning & Quality Assurance:
Handling Nulls: Check for sensor dropouts in rpm_variation or acceleration_smoothness. Use Linear Interpolation for time-series gaps or Iterative Imputer for cross-sectional gaps.

Outlier Detection: Use Z-score or Isolation Forest to identify non-physical sensor readings (e.g., impossible RPM spikes) that could skew eco-score correlations.

Feature Engineering (Microscopic Trajectory Analysis):
Kinetic Energy Recovery Index (KERI): Create a ratio of fuel_consumption to acceleration_smoothness to quantify energy lost during "stop-and-go" cycles.

Aggression Index: Combine harsh_braking_count and rpm_variation into a single weighted score to profile "aggressive" vs. "smooth" driving styles.

Idle Impact: Calculate idling_time as a percentage of total trip time to isolate fixed-cost fuel waste from signalized intersection delays.

Preprocessing for Modeling:
Feature Scaling: Apply StandardScaler to rpm_variation and fuel_consumption, as distance-based algorithms (like SVR) are sensitive to feature magnitude.

Multi-collinearity Check: Calculate Variance Inflation Factor (VIF). High correlation is expected between acceleration_smoothness and harsh_braking_count; removing redundant features improves model stability. 

II. Machine Learning Modeling Approach
For a dataset of this size with continuous targets, a Supervised Regression approach is recommended, followed by Unsupervised Profiling.

Regression Models (Predicting eco_score or fuel_consumption):
Baseline: Use Linear Regression to establish a performance floor.

Top Performer: Deploy Random Forest Regression (RFR). It handles non-linear relationships between "smoothness" and "fuel consumption" better than linear models, typically achieving R² scores > 0.95 in similar automotive datasets.

Alternative: XGBoost is highly effective for tabular telematics data and offers built-in handling of missing values.

Clustering (Unsupervised Driver Profiling):
Apply K-Means Clustering to segment drivers into categories: "Eco-Friendly," "Moderate," and "Aggressive". This allows for personalized real-time feedback systems rather than generic alerts.

Explainability (XAI):
Integrate SHAP (Shapley Additive Explanations) to show drivers why their score is low (e.g., "Your eco-score dropped 10% due to excessive idling at 5th Ave"). 

III. Actionable Strategies for 2026
Traffic Signal Optimization: Aggregate idling_time data by location to identify "hotspots" where signal synchronization can reduce stop-and-go cycles.
Real-time Feedback Loop: Deploy the trained RFR model into an edge device (e.g., connected car app) to provide instantaneous visual cues based on current acceleration_smoothness.


```{r}
#Data importation
library(readxl)
my_data <- read.csv("C:/Users/angel/OneDrive/Documents/DSA2050A DSCMETHODOLOGY/eco_driving_score.csv")

head(my_data) # Shows the first 6 rows
View(my_data) # Opens the data in a new RStudio tab


```
- There are 6 columns and 30,000 entries.
- The 6 columns represent the features of Eco-driving behavior of
most vehicle owners.

```{r}
# Checking for missing values
colSums(is.na(my_data))
```
- There are no missing values within the data set.
- It is already clean

```{r}
# Descriptive Analysis
str(my_data)
```
- There are 30,000 observations and 6 variables
- where they comprise of 2 variables input with integers and 4 variables input with numbers

```{r}
#Central Tendency
summary(my_data)
```
- The feature with the highest mean is rpm_variation of 1808. Which from the summary statistics shows the data is skewed to the right for the mean is greater than the median although they are close in value.
- The feature with the lowest mean is acceleration smoothness of 0.6677. Which from the summary statistics shows that it is skewed to the left for the mean is less than the median even though they are close in value.
-Because the difference between the mean and median is minimal the behavior of the vehicles is consistent and nothing outside the usual expectations can be detected.Meaning the outliers can be considered for futher speculation.
```{r}
# Variability / Dispersion
sd(my_data$rpm_variation)
```
- The sd of rpm_variation is low based by comparing the mean.
- sd of 678.424 and mean of 1808
- This indicates that the data points in the dataset are clustered tightly around the mean (average), signifying high consistency, low variability, and predictability.
```{r}
# Visualization for distribution
hist(my_data$rpm_variation, main="Distribtion of rpm variation", xlab="rpm_variation")


```
#Distribution of rpm Variation Interpretation
- Right-Skewed (Positive Skew): "The distribution of rpm Variation is right-skewed. The tallest bars are on the left (rpm variation), with a medium 'tail' of bars stretching to the right. This suggests that while most eco-driving behavior have a low to medium variation, it can be concluded that the cars run on diesel engines that have an rpm of 1,000-3,500 vs petrol for they have an rpm of 1,500-6,000 ". There for it can be concluded that most vehicles are diesel based on the range distribution in rpm variation.

- Concentration: "Approximately 80% of the data falls within the three central bins (1000–3000 rpm), indicating that there is fuel efficiency and environmental friendliness "

```{r}
hist(my_data$acceleration_smoothness, main="Distribtion of acceleration_smoothness", xlab="acceleration_smoothness")
```


#Distribution of acceleration_smoothness Interpretation
- left-Skewed (Negative Skew): "The distribution of acceleration_smoothness is left-skewed. The tallest bars are on the right (acceleration_smoothness), with a medium 'tail' of bars stretching to the left This suggests that most eco-driving behavior have a Linear Velocity Change: The speed increases in a predictable, linear, or fluid manner, it can be concluded that the cars have a lower mechanical wear and tear as well as fuel efficiency".

- Concentration: It can be seen that most vehicles have an acceleration_smoothness of 0.7 indicating the driver is applying gradual pressure to the accelerator/brake, leading to a comfortable, predictable, and fuel-efficient ride, as opposed to sudden jerks. In vehicle testing, this often relates to maintaining consistent, low-noise acceleration.

```{r}
hist(my_data$idling_time, main="Distribtion of idling_time", xlab="idling_time")
```


#Distribution of idling time Interpretation
- Right-Skewed (Positive Skew): "The distribution of idling time is right-skewed. The tallest bars are on the left (idling time), with a medium 'tail' of bars stretching to the right This suggests that most eco-driving behavior have an average idling time of 13 minutes due to traffic congestion, it can be concluded that the cars have a higher mechanical engine wear and tear as well as fuel wastage".

- Concentration: It can be seen that most vehicles have an idling time of 13 minutes. Therefore in conclusion there is Excessive Emissions: Idling for more than 10 seconds produces more pollutants and consumes more fuel than restarting the engine. 13 minutes of idling is equivalent to producing large amounts of CO2, a primary contributor to global warming.

```{r}
# Visualization to understand outliers
boxplot(my_data$rpm_variation, main='Boxplot of rpm_variation')

```
#Distribution of rpm Variation Interpretation
"The data is right-skewed; most vehicles are petrol based with an rpm raging from 1,800 - 3,800 rpm." There can also be seen that there are a few vehicles moving from 3,800 - 5,000 rpm indicating there is a potential that not all cars have diesel engines. 

```{r}
boxplot(my_data$acceleration_smoothness, main='Boxplot of acceleration_smoothness')
```

#Distribution of acceleration_smoothness Interpretation
"The data is left-skewed; most vehicles have an acceleration_smoothness below 0.7 indicating a high-efficiency driving style characterized by gentle, gradual acceleration and minimal sudden movements.Which means a smoother, more fuel-efficient approach that reduces fuel consumption. "

```{r}
boxplot(my_data$idling_time, main='Boxplot of idling_time')
```

#Distribution of idling time Interpretation
"The data is right-skewed; most vehicles have an idling time of 13 - 35 minutes indicating excessive and inefficient fuel consumption, representing a significant, preventable waste of fuel and money. "

```{r}
plot(my_data$rpm_variation, my_data$acceleration_smoothness, xlab = 'rpm_variation', ylab='acceleration_smoothness')
```
#Scatterplot conclusion
- Use statistical inference techniques to sample the data for it is too concentrated to interprate.

#Approach
To address the 15-25% increase in fuel waste and high CO2 emissions caused by aggressive driving and poor signal synchronization, the strategy is distilled into these core components:

3 Major Strategic Approaches
Behavioral Feature Engineering:
Transforming raw sensor data into Kinetic Energy Loss metrics. By correlating rpm_variation and acceleration_smoothness against fuel_consumption, the project quantifies exactly how much energy is "wasted" during stop-and-go cycles at intersections.

Predictive Eco-Efficiency Modeling:
Utilizing Ensemble Regressors (like Random Forest) to establish a baseline for "optimal" driving. This identifies the delta between current aggressive patterns and the theoretical minimum fuel consumption for specific urban routes, providing the mathematical foundation for the 15-25% reduction target.

Real-Time Feedback & Infrastructure Sync:
Using Unsupervised Clustering to profile driving styles and identify high-idle "hotspots." This dual approach allows for personalized driver alerts (Eco-driving feedback) and systemic urban planning (optimizing traffic signal timing to minimize idling_time).

1 Major Deployment Function
To deploy this in 2026, you would use a Real-Time Inference Wrapper. This function sits on an edge device (connected vehicle or traffic controller) and acts as the bridge between raw telemetry and actionable output.
